---
title: "parametra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{parametra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction to PARAMETRA

The PARAMETRA package provides access to a comprehensive database of parameters for livestock disease modeling. This vignette demonstrates how to install and use the package, explore the data, and perform some basic analyses.

## Parameters available 

The following table shows the current number of entries by pathogen and parameter type. To contribute new data, you can use the [PARAMETRA submission form](https://ec.europa.eu/eusurvey/runner/parametra-submission).

```{r packages, include = FALSE, warnings=FALSE, message=FALSE}
library(flextable)
library(scales)
library(tidyr)
library(parametra)
library(dplyr)
```


```{r setup, include = FALSE}
# Function to create a heatmap visualization
parametra_heatmap <- function(x, y, data=parametra_long, order_by=NULL, filename = NULL, na.rm=FALSE) {
  # Create data summary
  data <- data %>%
    dplyr::filter(!grepl(";", .data[[x]])) %>%  # Filter for only individual items
    dplyr::group_by(across(all_of(c(x, y)))) %>%
    dplyr::summarise(n = n())

  # Remove NA values if specified
  if(na.rm){
    data <- data %>%
      dplyr::filter(!is.na(.data[[y]]))
  }
  
  # Order rows based on specified criteria or default to descending count
  if(is.null(order_by)){
    data <- dplyr::arrange(data, desc(n))
  }else{
    data <- dplyr::arrange(data, across(all_of(order_by)), desc(n))
  }

  # Create pivot wider matrix
  matrix <- tidyr::pivot_wider(
    data = data[!is.na(data$n), ],
    id_cols = all_of(x),
    names_from = all_of(y),
    values_from = n
  )

  # Convert to matrix format
  rnames <- matrix[[x]]
  matrix[[x]] <- NULL
  matrix <- as.matrix(matrix)
  row.names(matrix) <- rnames

  # Replace NA with 0
  matrix[is.na(matrix)] <- 0

  # Save to file if filename provided
  if (!is.null(filename)) {
    matrix_df <- as.data.frame(matrix)
    matrix_df[[x]] <- row.names(matrix)
    write.csv(matrix_df, file = filename, row.names = FALSE)
  }

  return(matrix)
}

# Function to capitalize all character columns in a dataframe
capitalize_df <- function(df) {
  df[] <- lapply(df, function(x) {
    if(is.character(x)) {
      sapply(x, function(str) {
        s <- strsplit(str, " ")[[1]]
        paste(toupper(substring(s, 1, 1)), 
              substring(s, 2),
              sep = "", collapse = " ")
      })
    } else {
      x
    }
  })
  return(df)
}

# Function to capitalize individual strings
capitalize_vector <- function(x) {
  if(is.character(x)) {
    s <- strsplit(x, " ")
    sapply(s, function(words) {
      paste(toupper(substring(words, 1, 1)), 
            substring(words, 2),
            sep = "", collapse = " ")
    })
  } else {
    x
  }
}

# Shared color palettes for consistent aesthetics
primary_colors <- c("#014063", "#0F9DA4", "#ED684C","#8c96c6","#EDEDED")
parametra_palette <- function(n) colorRampPalette(primary_colors)(n)
parametra_grad <- function(n) colorRampPalette(c("#e0ecf4","#8c96c6", "#88419d", "#810f7c"))(n)
```

```{r ft_heatmap, include = FALSE}
ft_heatmap<-function(x,y,y_by=NULL, order_by=NULL){
  # Generate heatmap matrix
  heatmap_matrix <- parametra_heatmap(x=x, y=y, order_by=order_by)
  
  # Convert matrix to dataframe and handle NA values
  heatmap_df <- as.data.frame(heatmap_matrix)
  heatmap_df[is.na(heatmap_df)] <- 0
  heatmap_df[[x]] <- rownames(heatmap_matrix)
  heatmap_df<-heatmap_df[order(heatmap_df[[x]]),]
  names(heatmap_df)[names(heatmap_df)=="NA"]<-"Not available"
  
  # Prepare headers table
  headers_table <- unique(parametra_long[c(y,y_by)]) %>%
    filter(!.data[[y]]=="Other") 
  
  if(!is.null(order_by)){
    headers_table<-headers_table%>%
      arrange(across(all_of(order_by)))
  }
  

  
  # Prepare matrix header
  matrix_header <- data.frame(
      col_keys = headers_table[[y]],
      line2 = headers_table[[y]],
      line3 = if(is.null(y_by)) headers_table[[y]] else headers_table[[y_by]]
    )
  
  
  # Clean parameter types
  matrix_header <- matrix_header %>%
    mutate(
      # Handle parameter type categorization
      line3 = if(!is.null(y_by)&&y_by == "parameter_type") {
        case_when(
          line3 == "Transmission" ~ "Transmission",
          line3 == "InfectiousLatentIncubatperiod" ~ "Time intervals",
          line3 == "PathogenSurvival" ~ "Survival",
          line3 == "DiagnosticTest" ~ "Tests",
          line3 == "RegionalPrevalence" ~ "Prevalence",
          line3 == "WithinHerdPrevalence" ~ "Prevalence",
          line3 == "Other" ~ "Other",
          TRUE ~ line3
        )
      } else line3,
      
      # Handle study type categorization
      line3 = if(is.null(y_by)&&y== "study_type") {
        case_when(
          is.na(line2) ~ "Not available",
          line2 %in% c("Experimental", "Experimental Stable", 
                      "Experimental Isolator", "Experimental Field", 
                      "Experimental Lab") ~ "Experimental",
          line2 %in% c("Review", "Descriptive Review", 
                      "Meta-analysis") ~ "Review",
          line2 %in% c("Field", "Observational", "Seroprevalence",
                      "Prevalence", "Retrospective", 
                      "Retrospective cohort") ~ "Field",
          TRUE ~ "Other"
        )
      } else line3
    )
  
  # Add "Other" row
  last_row_header <- data.frame(col_keys="Other", line2="Other", line3="Other")
  matrix_header <- rbind(matrix_header, last_row_header)
  
  # Add "Not available" row
  na_row_header <- data.frame(col_keys="Not available", line2="Not available", line3="Not available")
  matrix_header <- rbind(matrix_header, na_row_header)
  
  # Order by line3 and line 2
  matrix_header <- matrix_header %>%
    arrange(
      line3 == "Not available",  # FALSE comes before TRUE
      line3 == "Other", 
      line3,
      line2 == "Not available",
      line2 == "Other",
      line2
    )

  
  # Add first row header
  first_row_header <- data.frame(col_keys=x, line2=x, line3=x)
  matrix_header <- rbind(first_row_header, matrix_header)
  

  # Capitalize headers and column names
  # Fix column naming by replacing "/" with "_"
  matrix_header[["col_keys"]] <- gsub("/", " or ", matrix_header[["col_keys"]])


  # Update and check column alignment
  names(heatmap_df) <- gsub("/", " or ", names(heatmap_df))
  heatmap_df<-heatmap_df[intersect(matrix_header$col_keys,names(heatmap_df))] 
  matrix_header<-matrix_header[match(intersect(matrix_header$col_keys,names(heatmap_df)),matrix_header$col_keys),] 

  parametra_count_x <- parametra_long %>%
    group_by(across(all_of(c(x, "ref")))) %>%
    summarise(entries = n(), .groups = 'drop') %>%
    group_by(across(all_of(c( x)))) %>%
    summarise(
      n_entries = sum(entries),
      n_studies = n_distinct(ref),
      .groups = 'drop'
    )
  
  parametra_count_y <- parametra_long %>%
    mutate(across(all_of(y), ~ifelse(is.na(.), "Not available", .))) %>%
    group_by(across(all_of(c(y, "ref")))) %>%
    summarise(entries = n(), .groups = 'drop') %>%
    group_by(across(all_of(c( y)))) %>%
    summarise(
      n_entries = sum(entries),
      n_studies = n_distinct(ref),
      .groups = 'drop'
    )
  
  if(is.null(y_by)&&y == "study_type") y_by<-"study_type2"

  if(is.null(y_by)) y_by<-y

  parametra_count_y_by <- parametra_long  %>%
    mutate(
      parameter_type = if(y_by == "parameter_type") {
        case_when(
          parameter_type == "Transmission" ~ "Transmission",
          parameter_type == "InfectiousLatentIncubatperiod" ~ "Time intervals",
          parameter_type == "PathogenSurvival" ~ "Survival",
          parameter_type == "DiagnosticTest" ~ "Tests",
          parameter_type == "RegionalPrevalence" ~ "Prevalence",
          parameter_type == "WithinHerdPrevalence" ~ "Prevalence",
          TRUE ~ parameter_type
        )
      } else parameter_type,
      
      study_type2=if(y_by=="study_type2") {
        case_when(
          is.na(study_type) ~ "Not available",
          study_type %in% c("Experimental", "Experimental Stable", 
                      "Experimental Isolator", "Experimental Field", 
                      "Experimental Lab") ~ "Experimental",
          study_type %in% c("Review", "Descriptive Review", 
                      "Meta-analysis") ~ "Review",
          study_type %in% c("Field", "Observational", "Seroprevalence",
                      "Prevalence", "Retrospective", 
                      "Retrospective cohort") ~ "Field",
          TRUE ~ "Other"
        )
      } else NULL)%>%
    mutate(across(all_of(y_by), ~ifelse(is.na(.), "Not available", .))) %>%
    group_by(across(all_of(c(y_by, "ref")))) %>%
    summarise(entries = n(), .groups = 'drop')%>%
    group_by(across(all_of(c( y_by)))) %>%
    summarise(
      n_entries = sum(entries),
      n_studies = n_distinct(ref),
      .groups = 'drop'
    )
  
  
  heatmap_df[[1]]<-paste0(heatmap_df[[1]]," (",parametra_count_x$n_studies[match(heatmap_df[[1]],parametra_count_x[[1]])],")")
  
  matrix_header$line2<-paste0(matrix_header$line2," (",parametra_count_y$n_studies[match(matrix_header$line2,parametra_count_y[[y]])],")")
  matrix_header$line2 <- gsub(" \\(NA\\)", "", matrix_header$line2)
  matrix_header$line2[1]<-""
  
  matrix_header$line3<-paste0(matrix_header$line3," (",parametra_count_y_by$n_studies[match(matrix_header$line3,parametra_count_y_by[[y_by]])],")")
  matrix_header$line3 <- gsub(" \\(NA\\)", "", matrix_header$line3)
  matrix_header$line3 <- ifelse(matrix_header$line3=="Other",matrix_header$line2,matrix_header$line3)
  matrix_header$line3[1]<-""
  
  matrix_header <- capitalize_df(matrix_header)
  names(heatmap_df) <- capitalize_vector(names(heatmap_df))
  
  # Color scale function
  color_scale <- function(x) {
    # Handle case where all values are 0
    if(all(x == 0, na.rm = TRUE)) return(rep("white", length(x)))
    
    # Replace NA with 0 and create base colors
    x[is.na(x)] <- 0
    colors <- rep("white", length(x))
    non_zero <- x > 0
    
    # Get all values from the entire table for global scaling
    all_values <- unlist(heatmap_df[,-1])  # Exclude first column (names)
    global_min <- min(all_values[all_values > 0], na.rm = TRUE)
    global_max <- max(all_values, na.rm = TRUE)
    
    if(any(non_zero)) {
      non_zero_x <- x[non_zero]
      # Scale based on global min and max
      scaled_values <- (non_zero_x - global_min) / (global_max - global_min)
      # Ensure no NA values in color assignment
      color_indices <- pmax(1, pmin(100, round(scaled_values * 99) + 1))
      colors[non_zero] <- parametra_grad(100)[color_indices]
    }
    
    return(colors)
  }
  
  # Create flextable with improved formatting
  ft <- flextable(heatmap_df, col_keys = matrix_header$col_keys) %>%
    set_header_df(mapping = matrix_header, key = "col_keys") %>%
    merge_h(part = "header") %>%
    merge_v(part = "header") %>%
    theme_vanilla() %>%
    rotate(j = 2:ncol(heatmap_df), rotation = "btlr", part = "header", i = 1) %>%
    align(align = "center", part = "header") %>%
    bg(bg = function(x) {
         x <- as.numeric(as.character(x))
         x[is.na(x)] <- 0
         colors <- color_scale(x)
         if(any(is.na(colors))) colors[is.na(colors)] <- "white"
         colors
       },
       j = 2:ncol(heatmap_df),
       part = "body") %>%
    autofit() %>%
    fontsize(size = 9, part = "all") %>%
    border_outer() %>%
    border_inner()
  
  ft <- set_caption(ft, caption = paste0("Number of parametra entries by ", x, " and ", y, ". The numbers in parentheses indicate the count of studies where these entries were collected. Total number of studies: ", length(unique(parametra_long$ref))))
  return(ft)
  
}

```


```{r heatmap_render, echo=FALSE, message=FALSE}
x <- "pathogen"
y <- "parameter"
y_by<-"parameter_type"
ft_heatmap(x,y,y_by)
```

## Using PARAMETRA

### Installation

```{r install, eval=FALSE}
# Install the package from GitHub
# install.packages("devtools")
devtools::install_github("BIOSECURE-EU/parametra")
```


```{r load_pkgs, warnings=FALSE, message=FALSE}
# Load packages
library(parametra)
library(dplyr) # Data manipulation
library(ggplot2) # Plots
library(ggiraph) # Interactive tooltips
```

### Database Overview

PARAMETRA contains multiple tables with different parameter types. Using the following code you can get an overview of what's available:

```{r db_overview, include=FALSE}
# Look at the available datasets
data(package = "parametra")

# Preview the main dataset
head(parametra_long)

# Summary of parameter types
table(parametra_long$parameter_type)

# Summary of pathogens
table(parametra_long$pathogen)
```


### Example 1: Exploring Transmission Parameters for a Specific Disease

Let's examine transmission parameters for a specific disease - CHANGE THIS TO SOMETHING COOL WITH FILTERS

```{r asf_transmission}
# Filter for African Swine Fever (ASF) transmission parameters
asf_transmission <- parametra_long %>%
  filter(pathogen == "African Swine Fever",
         parameter_type == "Transmission")

# View summary
summary(asf_transmission$value)

# Plot distribution of values
g <- ggplot(asf_transmission, aes(x = value, y = host)) +
  geom_point_interactive(
    aes(tooltip = paste("Value:",paste(value, ifelse(!is.na(lower_cl),paste0("(95% CI: ",lower_cl,"-", upper_cl,")"),"")),
                        "<br>Source:", ref_short, 
                        "<br>DOI:", ref,
                        "<br>Host:", host,
                        "<br>Study:", study_type,
                        ifelse(!is.na(if_other),paste("<br>Other:", if_other),""),
                        "<br>ID:", id),
        data_id = id),
    color = primary_colors[2],
    size = 3,
    alpha = 0.8,
    shape = 16,
    stroke = 0.3) +
  labs(x = "ASF transmission parameters",
       y = NULL) +
  theme_minimal(base_size = 12)+
  facet_wrap(~parameter,  ncol=1, scales="free")

# Convert to interactive graphic
girafe(ggobj = g, width_svg = 8, height_svg = 6)
```

### Example 2: Comparing Infectious Periods Across Pathogens

```{r infectious_periods}
# Extract infectious period data for multiple pathogens
infectious_periods <- parametra_long %>%
  filter(parameter == "Infectious period",
         pathogen %in% c("Foot and Mouth Disease",
                        "African Swine Fever", "Avian Influenza")) %>%
  filter(!is.na(value))

# Boxplot of infectious periods by pathogen with interactive individual points
inf_period__plot<-ggplot(infectious_periods, aes(x = pathogen, y = value)) +
  geom_boxplot(aes(fill = pathogen), size = 0.3, outlier.shape = NA, alpha = 0.4) +
  geom_point_interactive(
    aes(
      color = pathogen,
      tooltip = paste0("Value: ", value,
                       "<br>Host: ", ifelse(!is.na(host), host, ""),
                       "<br>Source: ", ifelse(!is.na(ref_short), ref_short, ""),
                       "<br>ID: ", id),
      data_id = id
    ),
    position = position_jitter(width = 0.2, height = 0),
    size = 1.6,
    alpha = 0.75,
    shape = 16
  ) +
  scale_fill_manual(values = primary_colors) +
  scale_color_manual(values = primary_colors) +
  labs(title = "Comparison of Infectious Periods Across Pathogens",
    x = "Pathogen", 
    y = "Infectious Period (days)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
     legend.position = "none")

girafe(ggobj = inf_period__plot, width_svg = 7, height_svg = 6)
```


### Example 3: Sensitivity vs Specificity by test 

```{r diagnostic_analysis}
# Extract diagnostic test sensitivity and specificity
diagnostic_tests <- parametra_long %>%
  filter(parameter_type == "DiagnosticTest",
         parameter %in% c("Sensitivity", "Specificity")) %>%
  filter(!is.na(value)) %>%
  mutate(value = ifelse(value > 1, value / 100, value)) %>%
  arrange(test_type)

# Build paired sensitivity/specificity per test_type + pathogen + ref (average duplicates)
sens_spec <- diagnostic_tests %>%
  select(test_type, pathogen, ref, parameter, value) %>%
  mutate(value = as.numeric(value)) %>%
  group_by(test_type, pathogen, ref, parameter) %>%
  summarize(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = parameter, values_from = value) %>%
  # Ensure numeric columns
  mutate(
    Sensitivity = as.numeric(Sensitivity),
    Specificity = as.numeric(Specificity)
  ) %>%
  # Keep rows where at least one is present; if one missing, use the other as approximation
  filter(!is.na(Sensitivity) & !is.na(Specificity))

# Classify tests into broad categories for the legend
sens_spec <- sens_spec %>%
  mutate(test_class = case_when(
    grepl("pcr|rt-?pcr", tolower(test_type)) ~ "PCR",
    grepl("sequenc|ngs|sanger", tolower(test_type)) ~ "Sequencing",
    grepl("culture|broth|agar", tolower(test_type)) ~ "Culture",
    grepl("elisa|igg|igm|serolog", tolower(test_type)) ~ "Serological",
    TRUE ~ "Other"
  ))

# Prepare palette for classes
classes <- unique(sens_spec$test_class)
class_cols <- parametra_palette(length(classes))
names(class_cols) <- classes

# Interactive scatter: Specificity (x) vs Sensitivity (y)
test_ss_plot <- ggplot(sens_spec, aes(x = Specificity, y = Sensitivity, color = test_class)) +
  geom_point_interactive(
    aes(
      tooltip = paste0(
        "Test: ", test_type,
        "<br>Pathogen: ", pathogen,
        "<br>Ref: ", ref,
        "<br>Sensitivity: ", round(Sensitivity, 3),
        "<br>Specificity: ", round(Specificity, 3)
      ),
      data_id = paste(test_type, pathogen, ref)
    ),
    size = 3,
    alpha = 0.85
  ) +
  scale_color_manual(values = class_cols) +
  labs(title = "Sensitivity vs Specificity by Test",
       x = "Specificity",
       y = "Sensitivity",
       color = "Test class") +
  theme_minimal(base_size = 12)

girafe(ggobj = test_ss_plot, width_svg = 7, height_svg = 6)
```