---
title: "Getting started with PARAMETRA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PARAMETRA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

PARAMETRA is an R data package containing curated parameters for livestock disease modelling. The package includes one combined dataset, `parametra_long`, plus one dataset per parameter group.

## Installation

```{r install, eval = FALSE}
# Install from GitHub
# install.packages("remotes")
remotes::install_github("BIOSECURE-EU/parametra")
```

```{r setup, message = FALSE, warning = FALSE}
library(parametra)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Data included in the package

The main dataset is `parametra_long`, which stacks all parameter-group tables into a single analysis-ready table. The column `parameter_type` identifies the original parameter group.

A quick overview of the number of records by parameter group:

```{r parameter-type-summary}
parametra_long %>%
  count(parameter_type, sort = TRUE)
```

And the number of records by pathogen:

```{r pathogen-summary}
parametra_long %>%
  count(pathogen, sort = TRUE)
```

## Finding relevant records

Most analyses start by filtering `parametra_long`. For example, the code below finds African swine fever transmission records with an available numeric value.

```{r filter-example}
asf_transmission <- parametra_long %>%
  filter(
    pathogen == "African Swine Fever",
    parameter_type == "transmission",
    !is.na(value)
  )
```

Use `distinct()` to see which values are available before filtering:

```{r available-values}
parametra_long %>%
  distinct(parameter_type, parameter) %>%
  arrange(parameter_type, parameter) %>%
  head(20)
```

## Example 1: Transmission parameters for African swine fever

This example plots transmission-parameter values for African swine fever, faceted by parameter. The record `id` is kept in the plotting data so the source row can be traced back to PARAMETRA.

```{r asf-transmission-plot, fig.width = 8, fig.height = 6}
asf_transmission %>%
  ggplot(aes(x = value, y = host)) +
  geom_point(color = "#0F9DA4", size = 2.5, alpha = 0.8) +
  facet_wrap(~ parameter, ncol = 1, scales = "free_x") +
  labs(
    title = "African swine fever transmission parameters",
    subtitle = "Each point is one PARAMETRA record",
    x = "Value",
    y = "Host"
  ) +
  theme_minimal()
```

## Example 2: Comparing infectious periods across pathogens

Here we compare infectious-period estimates for three pathogens. This example is useful for checking the range of values before selecting parameters for a model.

```{r infectious-periods, fig.width = 8, fig.height = 5}
infectious_periods <- parametra_long %>%
  filter(
    parameter == "Infectious period",
    pathogen %in% c("Foot and Mouth Disease", "African Swine Fever", "Avian Influenza"),
    !is.na(value)
  )
ggplot(infectious_periods, aes(x = pathogen, y = value, fill = pathogen)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.7, size = 1.5) +
  labs(
    title = "Infectious-period estimates across pathogens",
    x = NULL,
    y = "Infectious period"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Working with references

Each record includes reference fields so parameter values can be traced to their source. Useful columns include:

-   `ref`: DOI, DOI URL, PubMed URL, or stable web URL

-   `ref_short`: short human-readable citation, when available

-   `ref_status`: status assigned during PARAMETRA curation

-   `ref_last_access`: date when a URL reference was last checked.

```{r references}
parametra_long %>%
  count(ref_status, sort = TRUE)
```

## Interpreting PARAMETRA data

PARAMETRA is curated to support disease-modelling work, but users should still assess whether each record is appropriate for their specific modelling context. Before accepting a parameter value as suitable, we recommend reviewing the contextual information and notes provided in the database, and consulting the original reference.

## Contributing new records

New records can be submitted through the [PARAMETRA submission form](https://ec.europa.eu/eusurvey/runner/parametra-submission) or via [contact\@parametra.eu](mailto:contact@parametra.eu).
